\begin{abstract}

\begin{comment}
Our implementation of a verified instruction selection pipeline from a subset of LLVM IR, produced by the Lean compiler, to ARM machine code is a first step toward strengthening trust in Lean’s translation process and proof by reflection, with the long-term goal of a fully verified Lean theorem proving infrastructure. The Lean interactive theorem prover plays a crucial role in formal verification. While its kernel has been verified, its compiler remains largely unverified. This is concerning for proof by reflection, where computations act as certificates for proofs, making compiler correctness essential for trustworthiness. We counteract this gap by implementing and verifying an instruction selection pass that translates LLVM IR to ARM assembly in Lean4 while preserving program semantics. We target a subset of LLVM IR commonly produced by the Lean compiler—load/store operations and integer arithmetic— and translate it to an ISA exhibiting formal correctness guarantees. Our implementation leverages formal models of LLVM and ARM and incorporates peephole optimization patterns for improved instruction selection. By utilizing bv\_ decide, a tactic within Lean allowing for automated bit vector reasoning, we gain significantly more proof automation compared to other projects like CompCert. The primary results of this work establish that any translation from LLVM IR to ARM machine code using our pipeline provides formal correctness guarantees proven in Lean.   
 
\end{comment}

 
\end{abstract}
